---
title: "Circadian Phase Estimation of Depmap Melanoma Cell Line Samples"
author: "Theo Killian"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    nature:
      countIncrementalSlides: false
      slideNumberFormat: "%current% / %total%"
---

```{r setup, include=FALSE}
# Load libraries & set chunk options
suppressPackageStartupMessages(library("dplyr"))
library("tidyr")
library("tibble")
library("ggplot2")
library("viridis")
library("stringr")
library("ggrepel")
library("ggpubr")
library("readxl")
library("depmap")
library("ExperimentHub")
library("pheatmap")
library("DT")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
# Custom color palette
my_cols <- c("#0D0887FF", "#6A00A8FF", "#B12A90FF", "#E16462FF", "#FCA636FF", "#F0F921FF")
```

--

## Introduction

- In circadian biology, estimating the **circadian phase** of biological samples provides insight into the temporal regulation of gene expression. 
- Large-scale resources like **DepMap** often lack time-of-day annotations, making it difficult to detect rhythmic patterns.

**We demonstrate:**
- **Harmonic regression** with random pseudo-times (0–24h) → primarily fits noise.
- **Reference-based approach** using a time-annotated circadian dataset (**GSE161566**) to infer approximate phase.

---

```{r}
known_circadian_genes <- c(
  "RORA", "PER3", "PER1", "CRY2", "RORB", "PER2", "NR1D2", "RORC", "CRY1",
  "ARNTL", "NPAS2", "CLOCK", "NR1D1", "ARNTL2")

knitr::kable(known_circadian_genes)
```


---